15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 1

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
For instance, one mechanism not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @herms_emerald_2014; @fischer_disturbances_2013] (Fei et al 2014). Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

>>>>

The temperate forest biome plays a critical role in the global carbon cycle (C), accounting for almost half of the net global forest C sink [@harris_global_2021], and temperate deciduous forests alone sequester >300 Tg C yr^-1^ [@pugh_role_2019]. Although it is currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020]<!--check refs-->, and current global C models do not represent certain influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
For instance, the impact of nuisance species, is not represented in global C - i.e., the impact of indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, leading to detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
The impact of nuisance species on forest C budgets is also not considered in machine learning or niche models that seek to project future forest distribution and C stocks, nor is it in C offset projects [@wu_uncertainty_2023].
Nuisance species, including non-indigenous insect pests and pathogens and over-abundant herbivores, are dramatically impacting carbon cycling in temperate deciduous forests around the world, and failure to account for them significantly limits understanding of future forest health and C sequestration potential.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling and successional trajectories [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.
